The electric power grid has undergone a fundamental transformation over the last two
decades. The integration of digital communications, programmable logic controllers, and
IP-connected SCADA and Human-Machine Interface (HMI) systems has enabled
unprecedented operational efficiency but has simultaneously expanded the cyber attack
surface. High-profile incidents—most notably the Stuxnet worm \cite{langner2011stuxnet}
and the Ukrainian power grid attacks of 2015 and 2016—demonstrate that adversaries are
willing and able to traverse the IT/OT boundary and compromise energy infrastructure
\cite{nafees2023}.

Among the specific threats targeting these environments, malware remains one of the most
pervasive and damaging attack vectors. Malware encompasses a wide spectrum of malicious
software—from early viruses to modern ransomware-as-a-service (RaaS) platforms and
polymorphic code—designed to disrupt operations, exfiltrate data, or establish persistent
footholds across IT/OT boundaries \cite{maniriho2024}. Traditional detection is
broadly split into \emph{static analysis}, which inspects binary artefacts for signatures
and control-flow structures without execution, and \emph{dynamic analysis}, which
observes runtime behaviour in controlled sandboxes \cite{ayub2023}. Static approaches
are undermined by obfuscation and packing; dynamic approaches impose sandboxing overhead
incompatible with continuous OT operations. Machine learning bridges this gap by learning
behavioural patterns directly from operational telemetry, without requiring signature
updates or execution isolation \cite{onal2025}.

Windows Event Logs have emerged as a particularly valuable ML data source: they record
process creation, privilege assignment, service installation, object access, and logon
events with millisecond timestamps, providing a fine-grained behavioural footprint across
every Windows host in a smart grid operations centre \cite{achmad2025}. Regulatory
frameworks reinforce this choice: NERC CIP-007 mandates event logging and review for
all applicable Bulk Electric System assets, and IEC~62351 specifies audit trail
generation for power systems communications, making Windows Event Log collection a
compliance requirement rather than an optional deployment decision.

Despite this regulatory mandate, Windows Event Logs have received comparatively little
attention as a detection surface for ICS-specific attacks. The research community has
focused on network-level anomalies (e.g., Modbus/TCP deviations \cite{goldenberg2013modbus})
or dedicated ICS protocols, leaving host-based event logs underexplored. Meanwhile,
general system-log anomaly detection has advanced significantly through deep learning
approaches such as DeepLog \cite{du2017deeplog} and LogAnomaly \cite{zhao2019loganomaly},
and through log-parsing methods like Drain \cite{he2017drain}. Graph-based approaches
\cite{zhen2025} and provenance-graph detectors \cite{han2020} have demonstrated that
capturing the \emph{relational structure} of events improves detection of multi-stage
attacks, but impose computational overhead incompatible with real-time OT constraints.
None of these techniques have been systematically applied to Windows Event Logs within the
ICS/SCADA context.

This paper makes the following contributions:
\begin{itemize}
  \item A synthetic Windows Event Log generator that models four ICS-relevant attack
        patterns within a realistic smart grid topology.
  \item A session-level feature engineering pipeline tailored to the temporal structure of
        Windows Event Logs.
  \item A baseline evaluation of Random Forest classification under stratified
        cross-validation, establishing measurable success criteria.
  \item An open-source framework enabling reproducible ICS security research without
        access to operational data.
\end{itemize}

Section~\ref{sec:lit} reviews related work. Section~\ref{sec:proposed} states the central
hypothesis and research objectives. Section~\ref{sec:method} describes the methodology.
Section~\ref{sec:results} reports preliminary results, and Section~\ref{sec:conc} concludes.
