\label{sec:lit}

\subsection{Windows Event Log and Sysmon-Based Detection}

Windows Event Logs contain semantically rich, structured records of system activity.
He~\textit{et al.}\ \cite{he2017drain} introduced Drain, an online log-parsing tree that
maps raw log messages to templates in $O(\text{log}~n)$ time. Du~\textit{et al.}\
\cite{du2017deeplog} modelled log key sequences as a workflow using LSTM networks, detecting
anomalies as deviations from learned sequential patterns. Zhao~\textit{et al.}\
\cite{zhao2019loganomaly} extended this to LogAnomaly, capturing both sequential and
quantitative anomalies. Achmad~\textit{et al.}\ \cite{achmad2025} leveraged Sysmon-extended
Windows logs with supervised and unsupervised ML classifiers, achieving F1~=~0.99 on
IT-environment datasets. \"{O}nal and G\"{u}ven \cite{onal2025} systematically catalogued
novel Windows event types relevant to dynamic malware behaviour and demonstrated that
richer event coverage improves classifier accuracy (ROC-AUC 0.962$\pm$0.009). Collectively,
these works confirm Windows logs as a viable ML data source but leave ICS-specific event
semantics and OT temporal constraints unexplored.

\subsection{Graph-Based and Provenance-Aware Detection}

In response to limitations of isolated feature representations, graph-based methods
treat system events as interconnected entities rather than independent records. Zhen~\textit{et al.}\
\cite{zhen2025} transform audit logs into heterogeneous graphs and apply Graph Neural
Networks (GNNs) to capture structural dependencies among processes, files, and network
endpoints, achieving high detection rates on multi-stage attack scenarios. Han~\textit{et al.}\
\cite{han2020} proposed Unicorn, a provenance-graph detector using sketch-based online
summarisation for runtime detection of Advanced Persistent Threats. The spatial awareness
of these methods enables detection of complex attack chains invisible to feature-based
classifiers. However, graph construction and traversal impose computational overhead and
latency incompatible with real-time OT constraints.

\subsection{Hybrid Static-Dynamic Analysis}

Ayub~\textit{et al.}\ \cite{ayub2023} proposed RWArmor, combining static ML models with
dynamic behavioural analysis for early ransomware detection; empirical results demonstrate
92--97\% accuracy within 30--120 second observation windows. Anderson and Roth
\cite{anderson2018} released EMBER, an open dataset for training static PE malware
classifiers, establishing a reproducible benchmark for feature-engineering studies. These
hybrid approaches improve early detection fidelity but depend on sandbox infrastructure
unavailable in operational OT deployments.

\subsection{ICS/SCADA Intrusion Detection}

Industrial control system security demands domain-adapted detection strategies.
Nafees~\textit{et al.}\ \cite{nafees2023} surveyed cyber-physical situational awareness
in smart grids, identifying strict temporal bounds, operational sensitivity, and
IT/OT interdependence as the primary constraints distinguishing OT security from IT
security. Carcano~\textit{et al.}\ \cite{carcano2011scada} proposed multidimensional
critical-state analysis for SCADA, detecting command-sequence deviations. Goldenberg and
Wool \cite{goldenberg2013modbus} built deterministic Modbus/TCP models with zero false
positives. Mitchell and Chen \cite{mitchell2014survey} surveyed CPS intrusion detection,
noting that host-level methods remain rare. Goh~\textit{et al.}\ \cite{goh2016swat}
released the SWaT dataset for process-level anomaly evaluation. Zhu~\textit{et al.}\
\cite{zhu2011scada} formalised a SCADA attack taxonomy across all Purdue model levels.
Across this body of work, Windows host-log analysis as a detection layer within the ICS
hierarchy is conspicuously absent.

\subsection{AI/ML Methods for Intrusion Detection}

Liao~\textit{et al.}\ \cite{liao2013ids} reviewed supervised, unsupervised, and hybrid
methods, noting that ensemble methods consistently outperform single classifiers.
Breiman's Random Forest \cite{breiman2001} remains a strong baseline due to variance
reduction via bagging and native class-imbalance compensation. Chen and Guestrin's
XGBoost \cite{chen2016xgboost} extends gradient boosting to large-scale tabular problems.
Chen~\textit{et al.}\ \cite{chen2022} combined ML, NLP, and graph analytics for
proactive threat hunting, addressing the ``weak signal'' problem of isolated log events.
Tavallaee~\textit{et al.}\ \cite{tavallaee2009kdd} revealed severe redundancy in
standard benchmarks, motivating domain-specific synthetic datasets. Sommer and Paxson
\cite{sommer2010outside} argued that closed-world ML assumptions undermine operational
deployment, motivating our evaluation protocol.

\subsection{Research Gap}

Table~\ref{tab:related} summarises representative works and their key limitations.
A consistent pattern emerges: feature-based methods offer efficiency but treat events
as independent signals, missing multi-event attack narratives. Graph-based models
capture relational structure but incur latency and preprocessing costs incompatible
with OT real-time constraints. Hybrid approaches improve fidelity but require sandboxes.
All existing solutions are fundamentally \emph{IT-centric}: they assume latency is
negotiable, resources are abundant, and offline analysis is acceptable.

Smart grid control systems operate under a different reality: decisions must be made
within strict temporal bounds, systems cannot be paused or sandboxed, and any detection
mechanism must function within tight computational budgets. The result is a critical gap
between model sophistication and operational feasibility that this work addresses.

\begin{table}[!t]
  \caption{Summary of Representative Related Works}
  \label{tab:related}
  \centering
  \footnotesize
  \begin{tabular}{llll}
    \hline
    Reference & Method & Key Limitation & Env. \\
    \hline
    Achmad et al.\ \cite{achmad2025}   & Sysmon + ML      & No OT context      & IT \\
    \"{O}nal \& G\"{u}ven \cite{onal2025} & Windows events + ML & Sandbox dependency & IT \\
    Zhen et al.\ \cite{zhen2025}       & GNN on audit logs & High latency        & IT \\
    Han et al.\ \cite{han2020}         & Provenance graphs & Scalability limits  & IT \\
    Ayub et al.\ \cite{ayub2023}       & RWArmor hybrid   & Sandbox bottleneck  & IT \\
    Chen et al.\ \cite{chen2022}       & ML+NLP+Graph     & Synthetic data only & IT \\
    Nafees et al.\ \cite{nafees2023}   & Smart grid review & No implementation  & OT \\
    \hline
    \textbf{This work}                 & RF on WEL sessions & Synthetic data only & OT \\
    \hline
  \end{tabular}
\end{table}
